# Module 3 — Two Gradients at Once: Latin Squares and Row–Column Designs

> **Sub-course: Field Experiments in Agriculture & Plant Breeding**
> [← Module 2](02-crd-and-rcbd.md) · [Sub-course home](README.md) · [Next: Module 4 — Factorials, Split-Plots and Strip-Plots →](04-factorial-split-strip.md)

---

## 🧭 Learning Objectives

After this module you will be able to:

1. Explain when **two-way blocking** (rows *and* columns) is worth its cost.
2. Generate and map a Latin square (`agricolae`, `FielDHub`).
3. Analyse a Latin square and compare it with one-way blocking using real data.
4. Recognize resolvable **row–column designs** as the generalization for many entries.

---

## 🎯 The Big Picture

Module 1 showed that fields vary with direction. Often you do not know which direction
dominates, or variation runs both ways (slope one way, an old hedge the other). A **Latin
square** blocks in both directions at once: each treatment appears **once in every row and
once in every column**. Greenhouses (bench × distance from the wall), orchards and
glasshouse compartments are typical settings.

The price: a t × t Latin square needs exactly t replicates and loses 2(t − 1) degrees of
freedom to rows and columns. It works best for roughly 4–8 treatments. For more entries,
**row–column designs** keep the two-way control without requiring a square.

---

## 🧠 Core Intuition

| Design | Controls | Model |
|---|---|---|
| RCBD (rows as blocks) | one gradient | `y ~ row + trt` |
| Latin square | two gradients | `y ~ row + col + trt` |
| Resolvable row–column | two gradients, many entries, replicates as large blocks | `y ~ rep + trt + (1\|rep:row) + (1\|rep:col)` |

---

## 👁️ Visual Intuition — generating a Latin square

```{r lsd}
library(agricolae); library(desplot); library(agridat)
lsd <- design.lsd(trt = LETTERS[1:5], seed = 3, serie = 0)$book
names(lsd)[4] <- "trt"
head(lsd)
```

```{r lsd-map, fig.height = 3.6, out.width = "60%"}
desplot(lsd, trt ~ col * row, text = trt, cex = 1.2, show.key = FALSE,
        main = "5 x 5 Latin square: each letter once per row and column")
```

`FielDHub::latin_square()` produces the same kind of design with a field book, and
`FielDHub::row_column()` builds resolvable row–column designs for larger trials.

---

## 🔬 Worked Example — `goulden.latin`

A 5 × 5 Latin square with 5 treatments [@agridat_cran].

```{r goulden-map, fig.height = 3.6, out.width = "70%"}
g <- goulden.latin
desplot(g, yield ~ col * row, text = trt, cex = 1.2,
        main = "goulden.latin: yield, treatment letters")
```

We fit four models to the same data — ignoring structure, blocking on rows, blocking on
columns, and the full Latin-square model:

```{r goulden-models}
models <- c("yield ~ trt", "yield ~ factor(row) + trt", "yield ~ factor(col) + trt",
            "yield ~ factor(row) + factor(col) + trt")
tab <- do.call(rbind, lapply(models, function(f) {
  a <- anova(lm(as.formula(f), data = g))
  data.frame(model = f, residual_df = a["Residuals", "Df"],
             residual_MS = round(a["Residuals", "Mean Sq"], 2),
             p_treatment = signif(a["trt", "Pr(>F)"], 3))
}))
tab
```

- **Rows matter:** blocking on rows cuts the residual mean square from `r tab$residual_MS[1]`
  to `r tab$residual_MS[2]`.
- **Columns add little here:** the full Latin-square model reaches `r tab$residual_MS[4]`, at
  the cost of `r tab$residual_df[2] - tab$residual_df[4]` more error degrees of freedom.
- The Latin square was still a safe choice: before the trial, nobody knew which direction
  would matter. The analysis must follow the design, so report the full model.

```{r goulden-means}
library(emmeans)
fit_ls <- lm(yield ~ factor(row) + factor(col) + trt, data = g)
emmeans(fit_ls, ~ trt)
```

---

## ⚠️ Common Misconceptions

| ❌ The misconception | ✅ What is actually true — and what to do |
|---|---|
| **"Latin squares are always better than RCBDs."** | They cost degrees of freedom; with small squares (3 × 3) there is almost no error d.f. left. |
| **"Drop the columns from the model if they aren't significant."** | Analyse as designed; post hoc model reduction changes the meaning of the tests. |
| **"Latin squares work for 20 varieties."** | That needs 20 replicates; use row–column designs. |
| **"The square must be square in metres."** | It must be square in *number* of rows and columns. |

---

## 🧪 Spot the Flaw

> "Six irrigation treatments were applied in a 6 × 6 Latin square; irrigation lines run
> along the rows, so each row received one irrigation level throughout."

<details>
<summary>▶ Diagnosis</summary>

If a factor is applied to whole rows, it is not randomized within rows; it is confounded with
row. A Latin square requires each treatment once per row and per column. Hard-to-change
factors such as irrigation need a split-plot or strip-plot design (Module 4).
</details>

---

## 🔎 The Reviewer's Perspective

- **"Were both row and column effects in the analysis?"**
- **"Were there enough error degrees of freedom (t ≥ 4–5)?"**
- **"Was the randomization a random Latin square, not a standard cyclic one?"**

---

## 🛠️ Design Challenge

A glasshouse has 4 benches (north to south) and plants are placed at 4 distances from the
heating pipes. Test 4 nutrient solutions. Generate and map a design.

<details>
<summary>▶ Model solution</summary>

```{r challenge, fig.height = 3.4, out.width = "60%"}
gh <- design.lsd(trt = c("N1", "N2", "N3", "N4"), seed = 42, serie = 0)$book
names(gh)[4] <- "solution"
desplot(gh, solution ~ col * row, text = solution, cex = 1.1, show.key = FALSE,
        main = "Rows = benches, columns = distance from pipes")
```

A 4 × 4 square leaves (4 − 1)(4 − 2) = 6 error d.f. — little. Run two independent squares
(e.g. two compartments or two runs in time) and analyse them together, with square as an
additional factor.
</details>

---

## 🧑‍💻 Code-along Exercises

1. Analyse `agridat::cochran.latin` (`diff ~ row + col + operator`). Which blocking direction matters?
2. Generate a resolvable row–column design for 30 entries in 3 replicates with `FielDHub::row_column()` and map it.
3. Compute the relative efficiency of the Latin square versus an RCBD on rows for `goulden.latin`.

---

## ✅ Check Your Understanding

**⭐ Q1.** What property defines a Latin square?

**⭐ Q2.** How many error degrees of freedom does a 6 × 6 Latin square have?

**⭐⭐ Q3.** In `goulden.latin`, which blocking direction mattered, and how do you know?

**⭐⭐ Q4.** Why not use a Latin square for 25 breeding lines?

**⭐⭐⭐ Q5.** Propose a design for 40 lines in 2 replicates in a field with gradients in both directions.

---

## 📝 Sample Answers & Assessment

<details>
<summary>▶ Show sample answers</summary>

> **Q1 — Sample answer:** *"Each treatment once in each row and each column."* — **✔ 10/10.**

> **Q2 — Sample answer:** *"(6 − 1)(6 − 2) = 20."* — **✔ 10/10.**

> **Q3 — Sample answer:** *"Rows — the residual MS fell from `r tab$residual_MS[1]` to `r tab$residual_MS[2]`."* — **✔ 10/10.**

> **Q4 — Sample answer:** *"It would need 25 reps."* — **✔ 10/10.**

> **Q5 — Sample answer:** *"A resolvable row–column design; each replicate is a block, with rows and columns within replicates as incomplete blocks."* — **✔ 10/10.** Add a spatial analysis as a complement (Module 7).

**Rubric:** two-way designs must be justified by two-way variation and analysed with both blocking factors.
</details>

---

## 🧾 Module Summary

| Concept | One-line takeaway |
|---|---|
| **Latin square** | Each treatment once per row and column; two-way blocking. |
| **Cost** | t replicates required; 2(t − 1) d.f. spent on blocking. |
| **Analysis** | `y ~ row + col + trt`, analysed as designed. |
| **Scaling up** | Row–column designs for many entries. |

### 📇 Field Design Card — rows for this module

| Field | Your answer |
|---|---|
| Gradients in one or two directions? | |
| Design (RCBD/Latin square/row–column) | |
| Error d.f. available | |

---

## 🔗 Go Deeper

- Main course: [Ch. 5 — Blocking](../../chapters/05-blocking-and-batches.md)
- Incomplete-block and row–column designs: Module 5; [@piepho2006]

<!-- REFS -->

---

[← Module 2](02-crd-and-rcbd.md) · [Sub-course home](README.md) · [Next: Module 4 — Factorials, Split-Plots and Strip-Plots →](04-factorial-split-strip.md)
