# Module 8 — Multi-Environment Trials and G × E

> **Sub-course: Field Experiments in Agriculture & Plant Breeding**
> [← Module 7](07-spatial-analysis.md) · [Sub-course home](README.md) · [Next: Module 9 — Planning a Field Trial →](09-planning-field-trials.md)

---

## 🧭 Learning Objectives

After this module you will be able to:

1. Map and summarize a multi-environment trial (MET) with faceted `desplot` figures.
2. Fit a mixed model that separates genotype, G × E, and within-trial design effects.
3. Use variance components to decide between **more replicates** and **more locations**.
4. Explain why the environment, not the plot, is the replicate for recommendations.

---

## 🎯 The Big Picture

Varieties are released for a *region* and for *future seasons*, not for one field. Genotypes
often rank differently across environments — **genotype × environment interaction (G × E)**.
A single-site trial, however well designed, estimates the genotype effect *plus* its interaction
with that site and year. Multi-environment trials (METs) spread the testing across locations
and years, and mixed models combine them [@smith2005; @piepho1997].

---

## 🧠 Core Intuition

For a genotype mean over *L* locations with *R* replicates each:

$$\operatorname{Var}(\bar y_{g}) = \frac{\sigma^2_{GE}}{L} + \frac{\sigma^2_e}{L\,R}$$

- Extra **replicates** shrink only the plot-error term.
- Extra **locations** shrink both — including G × E, which replicates can never reduce.

So when G × E matters, **spread plots over more environments** rather than piling replicates
into one.

---

## 👁️ Visual Intuition — `besag.met`

64 genotypes, 6 counties (locations), 3 replicates each, incomplete blocks within replicates.

```{r met-map, fig.height = 6.5, fig.width = 9}
library(agridat); library(desplot)
b <- besag.met
desplot(b, yield ~ col * row | county, out1 = rep, out2 = block,
        out2.gpar = list(col = "grey40", lwd = 0.5), strip.cex = 1,
        main = "besag.met: six locations, replicates (thick) and incomplete blocks (thin)")
```

Locations differ in mean yield, and each field has its own spatial pattern.

```{r met-ge, fig.height = 4.2}
gm <- aggregate(yield ~ gen + county, data = b, FUN = mean)
top <- names(sort(tapply(gm$yield, gm$gen, mean), decreasing = TRUE))[1:6]
sub <- droplevels(subset(gm, gen %in% top))
interaction.plot(sub$county, sub$gen, sub$yield, type = "b", pch = 19, lwd = 2,
                 col = c("#2a78d6", "#eb6834", "#1baf7a", "#eda100", "#e87ba4", "#4a3aa7"),
                 xlab = "Location", ylab = "Mean yield", trace.label = "Genotype",
                 main = "G x E: the six best genotypes across locations")
```

Crossing lines mean rankings change between locations — G × E in action.

---

## 🔬 Worked Example — variance components and resource allocation

```{r met-model}
library(lme4)
fm <- lmer(yield ~ county + (1 | gen) + (1 | gen:county) +
             (1 | county:rep) + (1 | county:rep:block), data = b)
vc <- as.data.frame(VarCorr(fm))[, c("grp", "vcov")]
vc
s2g  <- vc$vcov[vc$grp == "gen"]
s2ge <- vc$vcov[vc$grp == "gen:county"]
s2e  <- vc$vcov[vc$grp == "Residual"]
```

Now compute the heritability of a genotype mean for different allocations of the same
number of plots:

```{r allocation}
h2 <- function(L, R, ge = s2ge) s2g / (s2g + ge / L + s2e / (L * R))
alloc <- data.frame(locations = c(1, 2, 3, 6), reps = c(6, 3, 2, 1))
alloc$plots_per_genotype <- alloc$locations * alloc$reps
alloc$H2_observed_GxE <- round(mapply(h2, alloc$locations, alloc$reps), 3)
alloc$H2_if_GxE_doubled <- round(mapply(h2, alloc$locations, alloc$reps, MoreArgs = list(ge = 2 * s2ge)), 3)
alloc
```

With the same 6 plots per genotype:

- **1 location × 6 replicates:** H² = `r alloc$H2_observed_GxE[1]`.
- **6 locations × 1 replicate:** H² = `r alloc$H2_observed_GxE[4]`.
- If G × E were twice as large, the advantage of spreading across locations would grow
  (`r alloc$H2_if_GxE_doubled[1]` vs `r alloc$H2_if_GxE_doubled[4]`).

In this dataset G × E is modest (σ²_GE = `r round(s2ge, 1)` vs σ²_G = `r round(s2g, 1)`), so the
difference is real but moderate. **Estimate these components from your own programme's
historical data before allocating plots.** Keep at least 2 replicates per location if you also
need location-specific results or within-site error estimates.

---

## ⚠️ Common Misconceptions

| ❌ The misconception | ✅ What is actually true — and what to do |
|---|---|
| **"More replicates at the station are as good as more sites."** | Not when G × E exists. |
| **"G × E is noise."** | It can be exploited (local adaptation) or managed (broad adaptation), but it must be estimated. |
| **"Analyse each site separately and average."** | A joint mixed model weights sites properly and estimates G × E. |
| **"One year is enough."** | Years are environments too; genotype × year interaction is often large. |

---

## 🧪 Spot the Flaw

> "To recommend a variety for the whole region, we tested 12 varieties at our research station
> with 8 replicates in one year."

<details>
<summary>▶ Diagnosis</summary>

One environment: genotype effects are confounded with genotype × (station, year) interaction.
96 plots per variety would serve better as e.g. 2 replicates at 4 locations over 2 years, with
a joint mixed-model analysis.
</details>

---

## 🔎 The Reviewer's Perspective

- **"How many locations and years, and how were they chosen (representative of the target region)?"**
- **"Was G × E estimated in a joint model?"**
- **"Is the recommendation's scope consistent with the environments tested?"**

---

## 🛠️ Design Challenge

You have budget for 360 plots to evaluate 30 maize hybrids. From past trials:
σ²_G = 40, σ²_GE = 30, σ²_e = 120. Find the allocation of locations × replicates with the highest
heritability, keeping at least 2 replicates per location.

<details>
<summary>▶ Model solution</summary>

```{r challenge}
h2c <- function(L, R) 40 / (40 + 30 / L + 120 / (L * R))
opts <- subset(expand.grid(L = 1:12, R = 2:6), L * R * 30 == 360)
opts$H2 <- round(h2c(opts$L, opts$R), 3)
opts[order(-opts$H2), ]
```

The most locations compatible with 2 replicates (6 × 2) maximizes heritability.
</details>

---

## 🧑‍💻 Code-along Exercises

1. Fit `besag.met` with a separate residual variance per location (`glmmTMB` or `nlme`). Do sites differ in precision?
2. Compute genotype BLUPs from the joint model and compare rankings with site-wise means.
3. Explore `agridat::belamkar.augmented` across 8 locations with faceted `desplot`.

---

## ✅ Check Your Understanding

**⭐ Q1.** What is G × E?

**⭐ Q2.** Which variance term can more replicates *not* reduce?

**⭐⭐ Q3.** Using the formula, explain why 6 locations × 1 rep beats 1 location × 6 reps.

**⭐⭐ Q4.** Why keep ≥ 2 replicates per location even if locations are more valuable?

**⭐⭐⭐ Q5.** How would you choose locations to represent a target region?

---

## 📝 Sample Answers & Assessment

<details>
<summary>▶ Show sample answers</summary>

> **Q1 — Sample answer:** *"When genotype rankings change between environments."* — **✔ 10/10.** (More generally: when genotype differences depend on the environment.)

> **Q2 — Sample answer:** *"G × E."* — **✔ 10/10.**

> **Q3 — Sample answer:** *"G × E is divided by 6 instead of 1."* — **✔ 10/10.**

> **Q4 — Sample answer:** *"To estimate within-site error and do site-level analyses."* — **✔ 10/10.**

> **Q5 — Sample answer:** *"Pick farms we know."* — **✘ 3/10.** Define the target population of environments (soils, climate zones, management), stratify the region, and sample locations within strata; include multiple years.

**Rubric:** link allocation to variance components and to the recommendation domain.
</details>

---

## 🧾 Module Summary

| Concept | One-line takeaway |
|---|---|
| **G × E** | Genotype differences depend on the environment. |
| **Allocation** | Var = σ²_GE/L + σ²_e/(LR): locations reduce both terms. |
| **Analysis** | Joint mixed model with genotype, G × E and design terms per site. |
| **Scope** | Recommend only for environments like those tested. |

### 📇 Field Design Card — rows for this module

| Field | Your answer |
|---|---|
| Target population of environments | |
| Locations × years × replicates | |
| Variance components used for planning (source) | |

---

## 🔗 Go Deeper

- Main course: [Ch. 19 — Breeding and Field Trials](../../chapters/19-breeding-and-field-trials.md) · [Ch. 9 — Descriptive Studies (sampling)](../../chapters/09-descriptive-studies.md)
- Reading: [@smith2005; @piepho1997; @crossa2017]

<!-- REFS -->

---

[← Module 7](07-spatial-analysis.md) · [Sub-course home](README.md) · [Next: Module 9 — Planning a Field Trial →](09-planning-field-trials.md)
