# Module 10 — Capstone: Design, Simulate, Analyse and Defend a Breeding Trial

> **Sub-course: Field Experiments in Agriculture & Plant Breeding**
> [← Module 9](09-planning-field-trials.md) · [Sub-course home](README.md)

---

## 🧭 Learning Objectives

After this capstone you will be able to:

1. Take a breeding trial from question to field book, at several locations.
2. **Simulate** a realistic trial (genetic effects, G × E, field trends) to test your design *before* sowing.
3. Compare analyses by how well they recover the **true** genetic values — something only simulation allows.
4. Write and defend a complete design dossier.

---

## 🎯 The Big Picture

In real trials you never know the true genetic values, so you can't check how good your design
and analysis really are. In a **simulated** trial you can. This capstone builds a complete trial
for **48 breeding lines at 3 locations**, generates yields with known truth, analyses them three
ways, and measures **selection accuracy**: the correlation between estimated and true genetic
values, and how many of the truly best lines you would have selected.

---

## 🔬 Part A — Worked example

### 1. Design: α-design, 2 replicates, blocks of 6, at 3 locations

```{r design}
library(agricolae); library(desplot)
make_site <- function(site, seed) {
  a <- design.alpha(trt = sprintf("L%02d", 1:48), k = 6, r = 2, seed = seed, serie = 0)$book
  names(a)[4] <- "gen"
  a$site <- site
  a$row  <- as.integer(as.character(a$cols))                # position within block (1-6)
  a$col  <- as.integer(as.character(a$block))               # blocks 1-16 side by side
  a
}
d <- rbind(make_site("S1", 11), make_site("S2", 12), make_site("S3", 13))
d$gen <- factor(d$gen); d$site <- factor(d$site)
d$rep <- factor(paste(d$site, d$replication)); d$blk <- factor(paste(d$site, d$block))
table(d$site)
```

```{r design-map, fig.height = 4.6, fig.width = 9}
desplot(d, replication ~ col * row | site, out1 = replication, out2 = block,
        out2.gpar = list(col = "grey30", lwd = 0.6), text = gen, cex = 0.45, shorten = "no",
        show.key = FALSE, main = "48 lines x 2 replicates x 3 sites (alpha-design, blocks of 6)")
```

### 2. Simulate yields with known truth

Each site gets a smooth trend **plus patchy, block-scale variation** (soil patches, wet spots),
G × E, and plot noise. The function returns one simulated dataset.

```{r simulate}
simulate_trial <- function(seed, d) {
  set.seed(seed)
  g  <- setNames(rnorm(48, 0, 4), levels(d$gen))                         # true genetic values (sd 4)
  ge <- setNames(rnorm(48 * 3, 0, 2.5), paste(rep(levels(d$gen), 3), rep(levels(d$site), each = 48)))
  site_mu <- c(S1 = 60, S2 = 52, S3 = 68)
  trend <- ifelse(d$site == "S1", 0.6 * d$col, ifelse(d$site == "S2", 4 * sin(d$col / 3), -0.4 * d$col + 1.5 * d$row))
  patch <- rnorm(nlevels(d$blk), 0, 4)[as.integer(d$blk)]                  # patchy field variation
  d$y <- site_mu[as.character(d$site)] + g[as.character(d$gen)] + ge[paste(d$gen, d$site)] +
         trend + patch + rnorm(nrow(d), 0, 3)
  list(data = d, truth = g)
}
one <- simulate_trial(2026, d)
```

```{r sim-map, fig.height = 4.2, fig.width = 9}
desplot(one$data, y ~ col * row | site, out1 = replication, main = "One simulated trial: trends and patches")
```

### 3. Analyse three ways and score them against the truth

```{r analyses}
library(lme4)
score <- function(sim) {
  dd <- sim$data; truth <- sim$truth[levels(dd$gen)]
  m_rep <- lmer(y ~ site + (1 | gen) + (1 | gen:site) + (1 | rep), data = dd)
  m_blk <- lmer(y ~ site + (1 | gen) + (1 | gen:site) + (1 | rep) + (1 | blk), data = dd)
  est <- list(naive_means    = tapply(dd$y, dd$gen, mean)[levels(dd$gen)],
              rep_only       = ranef(m_rep)$gen[levels(dd$gen), 1],
              rep_and_blocks = ranef(m_blk)$gen[levels(dd$gen), 1])
  top <- names(sort(truth, decreasing = TRUE))[1:8]
  c(sapply(est, cor, truth),
    sapply(est, function(e) sum(levels(dd$gen)[order(-e)][1:8] %in% top)))
}
res <- t(sapply(1:50, function(i) suppressMessages(score(simulate_trial(1000 + i, d)))))
summary_tab <- data.frame(analysis = c("naive means", "replicates only", "replicates + incomplete blocks"),
                          mean_accuracy = round(colMeans(res[, 1:3]), 3),
                          sd_accuracy = round(apply(res[, 1:3], 2, sd), 3),
                          mean_top8_recovered = round(colMeans(res[, 4:6]), 1))
summary_tab
```

```{r acc-plot, fig.height = 4}
boxplot(res[, 1:3], names = summary_tab$analysis, col = c("grey85", "#c9dcf5", "#2a78d6"),
        ylab = "Correlation with true genetic values", main = "Selection accuracy over 50 simulated trials")
```

```{r gain, echo = FALSE}
gain <- summary_tab$mean_accuracy[3] - summary_tab$mean_accuracy[1]
```

Over 50 simulated trials, the analysis that matches the design (replicates **and** incomplete
blocks) reaches a mean accuracy of `r summary_tab$mean_accuracy[3]` and recovers on average
`r summary_tab$mean_top8_recovered[3]` of the true top 8 lines, versus `r summary_tab$mean_accuracy[1]`
and `r summary_tab$mean_top8_recovered[1]` for naive means
(`r if (gain > 0.01) "a clear gain from modelling the design" else "little difference in this scenario"`).
The **spread** across simulated trials (SD ≈ `r summary_tab$sd_accuracy[3]`) is a reminder that any
single trial can be lucky or unlucky — which is why designs should be judged by simulation over many
runs, not one.

> **Try it:** set the patch SD in `simulate_trial()` to 0 (smooth fields only) and rerun. How much
> of the gain disappears? This shows *when* incomplete blocks pay off: when fields are patchy at
> the scale of the blocks.

---

## 🛠️ Part B — Your capstone project

### Brief

Choose **one** scenario (or your own):

1. **Early-generation wheat:** 300 lines, seed for ~1.3 plots per line, 1 site, strong east–west gradient.
2. **Advanced maize hybrids:** 40 hybrids, 4 locations × 2 years, budget 640 plots.
3. **Agronomy:** 3 tillage systems × 4 nitrogen rates × 2 cultivars; tillage machinery needs 20 m strips.
4. **Livestock forage:** 12 grass varieties grazed by sheep; grazing is managed per paddock.

Produce a dossier (R Markdown recommended) containing:

1. **Question and type** (main course, [Ch. 2](../../chapters/02-start-with-the-question.md)).
2. **Units, factors and strata** (which factor on which unit).
3. **Design choice** with justification (Modules 2–6) and the **field map** (`desplot`).
4. **Field book** generated with a recorded seed (`agricolae`/`FielDHub`).
5. **Planning**: replicates/locations from CV, Δ and variance components (Modules 8–9).
6. **Simulation** of the trial and analysis under realistic assumptions, reporting power or
   selection accuracy averaged over ≥ 100 simulated trials.
7. **Analysis plan**: model formula(s), spatial terms, how means/BLUPs and SEDs will be reported.
8. **Practicalities**: borders, plot size, data-quality rules, timeline.
9. **Risk register**: three things most likely to go wrong and mitigations.

### Rubric (100 points)

| Criterion | Full marks | Points |
|---|---|---|
| Question, type and scope match the design | precise; limits stated | 10 |
| Units and strata correct | each factor's unit and error stratum identified | 15 |
| Design justified and mapped | appropriate for entries, seed and gradients; clear `desplot` map | 15 |
| Reproducible field book | software + seed; coordinates included | 10 |
| Planning numbers | CV/variance components sourced; replicates/locations justified | 15 |
| Simulation | realistic, repeated, linked to the decision | 15 |
| Analysis plan | model matches design; spatial and MET terms where needed | 10 |
| Practicalities and risks | borders, data-quality rules, realistic risks | 10 |

### Peer review questions

1. Which factor is on which experimental unit? Is each tested against the right error?
2. Does the map show blocks oriented sensibly relative to the gradient?
3. Is the replication justified with numbers?
4. Would you trust the simulation's assumptions? What would change the conclusion most?

---

## 🧑‍💻 Code-along Exercises

1. Change the genetic SD (4) and G × E SD (2.5) in `simulate_trial()`. How does accuracy respond?
2. Add a SpATS spatial model per site (Module 7) and combine the site BLUEs in a second stage. Does accuracy improve?
3. Spend the same 288 plots as 1 replicate at 6 sites. Compare accuracy with the 2 reps × 3 sites design.

---

## 🧾 Sub-course Summary

| Module | Big idea |
|---|---|
| 1 | Measure field heterogeneity; block along gradients; Smith's law. |
| 2 | Randomize with seeds, map, analyse as designed. |
| 3 | Two-way blocking with Latin squares and row–column designs. |
| 4 | Split- and strip-plots have several error strata. |
| 5 | α-designs keep blocks small and recover information. |
| 6 | Augmented and p-rep designs for seed-limited trials. |
| 7 | Spatial models capture what blocks miss. |
| 8 | Environments, not plots, are the replicates for recommendations. |
| 9 | Plan replication from CV and Δ; simulate complex designs; keep field books and quality flags. |
| 10 | Simulate before sowing; judge designs by how well they recover the truth. |

---

## 🔗 Go Deeper

- Main course: [Ch. 26 — Capstone: The Design Clinic](../../chapters/26-capstone-design-clinic.md)
- Mixed models for breeding trials: [@smith2005; @piepho2003]

<!-- REFS -->

---

[← Module 9](09-planning-field-trials.md) · [Sub-course home](README.md)
